Template for indicator documentation

### 2023-09-14 DRAFT_NDVI_seminat.Rmd
starting the seminat markdown

# NDVI (greenness) indicator, wetland ecosystems  {#NDVI-indicator-wetland}

@@

June 2023

@@

## Introduction
The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e. national maps for the forest and mountain ecosystems). Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e. excluding forest swamps). For this task, two initial challenges can be identified early on:
1) We don't yet have a national wetland map, i.e. we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.

In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem. However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area. Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data. However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data. In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurments. We also still explore NDVI data from MODIS (see more details below).

## About the underlying data
In the NDVI project for wetlands, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:

- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
- remote sensed NDVI data
  - Sentinel-2 satellite NDVI data
  - MODIS satellite NDVI data
  - Landsat satellite NDVI data

@@

### Representativity in time and space
For wetlands, the NiN data contain 6502 wetland polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 6502 wetland polygons are distributed across regions and major ecosystem types in the following way:

@@

The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data. Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.

### Temporal coverage
The currently available NiN data on wetland ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. Note that none of the polygons in the NiN-data has been visited more than once during that period.

@@

## Collinearities with other indicators
The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in drained wetland areas under shrub and tree encroachment.

## Reference state and values
### Reference state and scaling values
This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for wetland ecosystems (cf. Jakobsson et al. 2020, Töpper & Jakobsson 2021). In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to wetland ecology and condition and (2) its applicability for defining a reference state. We will use statistical models to test whether or not NDVI varies systematically with condition and across time.

@@

## Analyses
### Data sets
The analyses in this document make use of the following data sets:
-NiN data (Naturtyper etter Miljødirektoratets innstruks)
-Sentinel-2 NDVI data
-MODIS NDVI data
-Landsat NDVI data

@@

## filter out only wetland data
nin.wetland <- nin %>% 
  filter(hovedoekosystem %in% c('Vaatmark')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
nin.wetland

colnames(nin.wetland)[c(1,9)] <- c("id","region_id")

# check that every nin-polygon still occurs only once
summary(as.factor(nin.wetland$id))
nin.wetland[nin.wetland$id=="NINFP2110008060",]
# this one was assigned to both Western and Central Norway, we drop the latter
nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% '1144.1',]

nin.wetland <- nin.wetland %>% 
  mutate(area_meters_nin = st_area(nin.wetland)
  )
# check and edit the order of regions
levels(nin.wetland$region)
nin.wetland$region <- as.factor(nin.wetland$region)
levels(nin.wetland$region)
nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
levels(nin.wetland$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.wetland2 <- nin.wetland %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
  separate(col = ninbeskrivelsesvariabler,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.wetland2 <- as.data.frame(nin.wetland2)

@@

#load processed NiN-data from cache
nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
nin.wetland2 <- readRDS(paste0(here::here(),"/data/cache/nin.wetland2.RDS"))

@@

#### NDVI data
```{r, results='hide'}
## Sentinel-2
# join nin.wetland & Sentinel NDVI data
SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
#summary(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.wetland))
#summary(SentinelNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
#summary(SentinelNDVI.wetland)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.wetland)
# split date into year, month & day
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

summary(SentinelNDVI.wetland)

## MODIS
# join nin.wetland and Modis NDVI data
ModisNDVI.wetland <- full_join(nin.wetland, df.m, by="id")

#summary(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(ModisNDVI.wetland))
#summary(ModisNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(hovedtype))
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
#summary(ModisNDVI.wetland)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.wetland)
# split date into year, month & day
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))
#summary(ModisNDVI.wetland)

## Landsat
# join nin.wetland & Landsat NDVI data
LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by="id")

#summary(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(LandsatNDVI.wetland))
#summary(LandsatNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(hovedtype))
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
#summary(LandsatNDVI.wetland)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.wetland)
# split date into year, month & day
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

#summary(LandsatNDVI.wetland)
```

@@

# How are polygon sizes distributed?
```{r}
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
abline(v=100,lty=2)
```
Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.

# How does NDVI vary over the years
```{r}
SentinelNDVI.wetland %>%
  ggplot( aes(x=year, y=mean )) + 
  geom_point() +
  facet_grid( tilstand~hovedtype)
```
2022 stands out with the highest NDVI values missing.
WE omit 2022 for this analysis
```{r}
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
```

# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=hovedtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey")

SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=subtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
```
NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3). The overlaps between the ecosystem types are considerable.
With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.

# How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x=region, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
```
There is no obvious trends in geography.

# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
```
In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern. Overlaps between condition classes are generally very large.

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r}
nin.wetland2 %>%
  filter(hovedtype=='V9') %>%
  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "1AG-B"="Shrub cover",
                                 "7GR-GI"="Abundance of ditches",
                                 "1AG-A-G"="Tree cover",
                                 "1AR-C-L"="Woody plants in field layer",
                                 "7TK"="Heavy vehicles",
                                 "PRSL"="Wear/tear")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition and condition variables in semi-natural mires (V9)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

>>>>

# NDVI (greenness) indicator, semi-natural ecosystems  {#NDVI-indicator-seminat}

@@

September 2023

@@

## Introduction
The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e. national maps for the forest and mountain ecosystems). Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems. For this task, two initial challenges can be identified early on:
1) We don't yet have a national map for semi-natural ecosystems, i.e. we don't know the exact locations of all semi-natural areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
2) The spatial extent of the single semi-natural ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.

In this draft development for a NDVI indicator for ecological condition in semi-natural  systems we approach these challenges in the following ways:
1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem. However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area. Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of semi-natural ecosystems are available.
2) We include Sentinel-2 imagery for semi-natural ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all semi-natural polygons in the NiN-data. However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data. In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurements. We also still explore NDVI data from MODIS (see more details below).

## About the underlying data
In the NDVI project for semi-natural systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:

- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets instruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
- remote sensed NDVI data
  - Sentinel-2 satellite NDVI data
  - MODIS satellite NDVI data
  - Landsat satellite NDVI data

@@

### Representativity in time and space
For semi-natural systems, the NiN data contain 6502 semi-natural polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 6502 semi-natural polygons are distributed across regions and major ecosystem types in the following way:

@@

The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data. Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the semi-natural occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.

### Temporal coverage
The currently available NiN data on semi-natural ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. Note that none of the polygons in the NiN-data has been visited more than once during that period.

@@

## Collinearities with other indicators
The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in abandoned semi-natural areas under shrub and tree encroachment.

## Reference state and values
### Reference state and scaling values
This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for semi-natural ecosystems (cf. Jakobsson et al. 2020, Töpper & Jakobsson 2021). In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to semi-natural ecology and condition and (2) its applicability for defining a reference state. We will use statistical models to test whether or not NDVI varies systematically with condition and across time.

@@

## Analyses
### Data sets
The analyses in this document make use of the following data sets:
-NiN data (Naturtyper etter Miljødirektoratets instruks)
-Sentinel-2 NDVI data
-MODIS NDVI data
-Landsat NDVI data

@@

## filter out only semi-natural data
nin.semi-natural <- nin %>% 
  filter(hovedoekosystem %in% c('Vaatmark')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.semi-natural = st_join(nin.semi-natural, regnor, left = TRUE)
nin.semi-natural

colnames(nin.semi-natural)[c(1,9)] <- c("id","region_id")

# check that every nin-polygon still occurs only once
summary(as.factor(nin.semi-natural$id))
nin.semi-natural[nin.semi-natural$id=="NINFP2110008060",]
# this one was assigned to both Western and Central Norway, we drop the latter
nin.semi-natural <- nin.semi-natural[!row.names(nin.semi-natural) %in% '1144.1',]

nin.semi-natural <- nin.semi-natural %>% 
  mutate(area_meters_nin = st_area(nin.semi-natural)
  )
# check and edit the order of regions
levels(nin.semi-natural$region)
nin.semi-natural$region <- as.factor(nin.semi-natural$region)
levels(nin.semi-natural$region)
nin.semi-natural$region <- factor(nin.semi-natural$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
levels(nin.semi-natural$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.semi-natural2 <- nin.semi-natural %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
  separate(col = ninbeskrivelsesvariabler,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.semi-natural2 <- as.data.frame(nin.semi-natural2)

@@

#load processed NiN-data from cache
nin.semi-natural <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural.RDS"))
nin.semi-natural2 <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural2.RDS"))

@@

#### NDVI data
```{r, results='hide'}
## Sentinel-2
# join nin.semi-natural & Sentinel NDVI data
SentinelNDVI.semi-natural <- full_join(nin.semi-natural, df.s, by="id")
#summary(SentinelNDVI.semi-natural)
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.semi-natural))
#summary(SentinelNDVI.semi-natural)
# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(hovedtype))
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(mean))
#summary(SentinelNDVI.semi-natural)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.semi-natural)
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.semi-natural)
# split date into year, month & day
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

summary(SentinelNDVI.semi-natural)

## MODIS
# join nin.semi-natural and Modis NDVI data
ModisNDVI.semi-natural <- full_join(nin.semi-natural, df.m, by="id")

#summary(ModisNDVI.semi-natural)
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(ModisNDVI.semi-natural))
#summary(ModisNDVI.semi-natural)
# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(hovedtype))
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(mean))
#summary(ModisNDVI.semi-natural)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.semi-natural)
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.semi-natural)
# split date into year, month & day
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))
#summary(ModisNDVI.semi-natural)

## Landsat
# join nin.semi-natural & Landsat NDVI data
LandsatNDVI.semi-natural <- full_join(nin.semi-natural, df.l, by="id")

#summary(LandsatNDVI.semi-natural)
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(LandsatNDVI.semi-natural))
#summary(LandsatNDVI.semi-natural)
# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(hovedtype))
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(mean))
#summary(LandsatNDVI.semi-natural)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.semi-natural)
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.semi-natural)
# split date into year, month & day
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

#summary(LandsatNDVI.semi-natural)
```

@@

# How are polygon sizes distributed?
```{r}
summary(SentinelNDVI.semi-natural$area_meters)
hist(SentinelNDVI.semi-natural$area_meters,xlim=c(0,20000),breaks=50000)
abline(v=100,lty=2)
```
semi-natural polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.

# How does NDVI vary over the years
```{r}
SentinelNDVI.semi-natural %>%
  ggplot( aes(x=year, y=mean )) + 
  geom_point() +
  facet_grid( tilstand~hovedtype)
```
2022 stands out with the highest NDVI values missing.
WE omit 2022 for this analysis
```{r}
SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(year != '2022')
```

# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.semi-natural %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=hovedtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey")

SentinelNDVI.semi-natural %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=subtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
```
NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3). The overlaps between the ecosystem types are considerable.
With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.

# How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.semi-natural %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x=region, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
```
There is no obvious trends in geography.

# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.semi-natural %>%
  filter(year == kartleggingsaar) %>%
  ggplot( aes(x=tilstand, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype)
```
In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern. Overlaps between condition classes are generally very large.

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r}
nin.semi-natural2 %>%
  filter(hovedtype=='V9') %>%
  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "1AG-B"="Shrub cover",
                                 "7GR-GI"="Abundance of ditches",
                                 "1AG-A-G"="Tree cover",
                                 "1AR-C-L"="Woody plants in field layer",
                                 "7TK"="Heavy vehicles",
                                 "PRSL"="Wear/tear")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition and condition variables in semi-natural mires (V9)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.semi-natural2 %>%
  filter(hovedtype=='V10') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
  theme(legend.position="none") +
  ggtitle("Condition and condition variables in semi-natural wet-meadows (V10)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
